---
title: "bayesDREAM: Cis variability across 100 seeds"
output:
  html_document:
    fig_width: 9
    fig_height: 5
    theme: minimal
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center")
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

# ── Path to output ─────────────────────────────────────────────────────────
OUT_DIR <- file.path(dirname(rstudioapi::getSourceEditorContext()$path),
                     "output_illustration")
if (!dir.exists(OUT_DIR)) OUT_DIR <- "output_illustration"

# ── Load and join data ─────────────────────────────────────────────────────
df   <- read.csv(file.path(OUT_DIR, "cis_variability_100seeds.csv"))
conv <- read.csv(file.path(OUT_DIR, "seed_convergence.csv"))
df   <- merge(df, conv[, c("seed", "converged", "final_loss")], by = "seed")

# ── Guide ordering (ascending true FC) and colours ────────────────────────
guide_order <- df %>%
  distinct(guide, true_fc) %>%
  arrange(true_fc) %>%
  pull(guide)

df$guide <- factor(df$guide, levels = guide_order)

guide_colours <- c(
  "KO1"   = "#1A237E",
  "KO2"   = "#3949AB",
  "KO3"   = "#90CAF9",
  "NTC_1" = "#2ECC71",
  "CA1"   = "#EF9A9A",
  "CA2"   = "#C62828"
)

conv_colours <- c("True" = "#2ECC71", "False" = "#E74C3C")
conv_shapes  <- c("True" = 16,        "False" = 4)

true_vals <- df %>% distinct(guide, true_fc)

theme_bd <- theme_classic(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    axis.title      = element_text(size = 11),
    legend.position = "none"
  )
```

## Convergence note

**97/100 converged** with niters=10 000 (tolerance 1e-4); 3 did not.
Convergence steps ranged from 692 to 9612.

---

## Plot 1: Estimated log₂FC per guide across 100 seeds

Each dot is one seed (● = converged, × = did not converge). Black diamond = true log₂FC.

```{r plot1-jitter, fig.height=4.5, fig.width=7}
ggplot(df, aes(x = guide, y = est_fc, colour = guide)) +
  geom_jitter(aes(shape = converged), alpha = 0.55, size = 1.8,
              position = position_jitter(width = 0.18, seed = 1)) +
  geom_point(data = true_vals, aes(x = guide, y = true_fc),
             shape = 18, size = 5, colour = "black", inherit.aes = FALSE) +
  geom_hline(yintercept = 0, colour = "grey70", linetype = "dashed",
             linewidth = 0.5) +
  scale_colour_manual(values = guide_colours) +
  labs(title = "Estimated log2FC across 100 seeds",
       x = "Guide", y = "Estimated log2FC") +
  theme_bw() +
  theme(legend.position = "bottom")
```

---

## Plot 1b: Per-cell x_true correlation by guide

For each seed × guide, Pearson and Spearman r between the latent true x (Normal draw)
and the model's per-cell posterior mean x_true, computed across cells within that guide.
Each dot is one seed (● = converged, × = did not converge).

```{r plot1b-xtrue-corr, fig.height=8, fig.width=7}
p_pearson <- ggplot(df, aes(x = guide, y = r_pearson, colour = guide)) +
  geom_jitter(aes(shape = converged), alpha = 0.55, size = 1.8,
              position = position_jitter(width = 0.18, seed = 1)) +
  geom_hline(yintercept = 0, colour = "grey70", linetype = "dashed", linewidth = 0.5) +
  scale_colour_manual(values = guide_colours) +
  labs(title = "Pearson r: true x vs estimated x_true",
       x = "Guide", y = "Pearson r") +
  theme_bw() +
  theme(legend.position = "bottom")

p_spearman <- ggplot(df, aes(x = guide, y = r_spearman, colour = guide)) +
  geom_jitter(aes(shape = converged), alpha = 0.55, size = 1.8,
              position = position_jitter(width = 0.18, seed = 1)) +
  geom_hline(yintercept = 0, colour = "grey70", linetype = "dashed", linewidth = 0.5) +
  scale_colour_manual(values = guide_colours) +
  labs(title = "Spearman r: true x vs estimated x_true",
       x = "Guide", y = "Spearman r") +
  theme_bw() +
  theme(legend.position = "bottom")

p_pearson / p_spearman

p_pearson <- ggplot(df, aes(x = guide, y = r_pearson, fill = guide)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, colour = "grey70", linetype = "dashed", linewidth = 0.5) +
  scale_fill_manual(values = guide_colours) +
  labs(title = "Pearson r: true x vs estimated x_true",
       x = "Guide", y = "Pearson r") +
  theme_bw() +
  theme(legend.position = "bottom")

p_spearman <- ggplot(df, aes(x = guide, y = r_spearman, fill = guide)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, colour = "grey70", linetype = "dashed", linewidth = 0.5) +
  scale_fill_manual(values = guide_colours) +
  labs(title = "Spearman r: true x vs estimated x_true",
       x = "Guide", y = "Spearman r") +
  theme_bw() +
  theme(legend.position = "bottom")

p_pearson / p_spearman
```

---

## Error summary

```{r error-summary}
df %>%
  group_by(guide) %>%
  summarise(
    true_fc   = first(true_fc),
    mean_est  = mean(est_fc),
    mean_bias = mean(bias),
    sd_bias   = sd(bias),
    pct_large = mean(abs(bias) > 0.3),
    .groups   = "drop"
  ) %>%
  arrange(true_fc) %>%
  mutate(across(where(is.numeric), \(x) round(x, 3)),
         pct_large = scales::percent(pct_large, accuracy = 1)) %>%
  knitr::kable(
    col.names = c("Guide", "True FC", "Mean est", "Mean bias", "SD bias", "|bias|>0.3"),
    align = "lrrrrr"
  )
```

---

## Plot 2: Pairwise correlation of bias across seeds

High positive correlation = the same seeds produce large errors in the same direction for both guides.

```{r plot2-pairwise-corr, fig.height=11, fig.width=12}
make_cor_plot <- function(data, title) {
  wide <- data %>%
    select(seed, guide, bias) %>%
    pivot_wider(names_from = guide, values_from = bias) %>%
    select(-seed)
  cm  <- cor(wide, method = "pearson")
  gns <- rownames(cm)
  cl  <- data.frame(
    guide1 = factor(rep(gns, times = length(gns)), levels = guide_order),
    guide2 = factor(rep(gns, each  = length(gns)), levels = rev(guide_order)),
    r      = as.vector(cm)
  )
  ggplot(cl, aes(x = guide1, y = guide2, fill = r)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    geom_text(aes(label = round(r, 2)), size = 3.5, colour = "black") +
    scale_fill_gradient2(low = "#3949AB", mid = "white", high = "#C62828",
                         midpoint = 0, limits = c(-1, 1), name = "Pearson r") +
    scale_x_discrete(position = "top") +
    labs(title = title, x = NULL, y = NULL) +
    theme_classic(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 12),
          axis.text = element_text(size = 10), axis.ticks = element_blank(),
          legend.position = "right")
}

p_all     <- make_cor_plot(df,
               paste0("All seeds (n=", n_distinct(df$seed), ")"))
p_conv    <- make_cor_plot(filter(df, converged == "True"),
               paste0("Converged (n=",
                      n_distinct(filter(df, converged == "True")$seed), ")"))
p_notconv <- make_cor_plot(filter(df, converged == "False"),
               paste0("Did not converge (n=",
                      n_distinct(filter(df, converged == "False")$seed), ")"))

p_all / (p_conv + p_notconv)
```

---

## Plot 3: Final loss vs absolute bias, facetted by guide

Does higher final ELBO loss predict larger estimation error?

```{r plot3-loss-vs-error, fig.height=6, fig.width=9}
ggplot(df, aes(x = final_loss, y = bias,
               colour = converged, shape = converged)) +
  geom_point(alpha = 0.6, size = 1.8) +
  facet_wrap(~ guide, scales = "free_x", nrow = 2) +
  scale_colour_manual(values = conv_colours, name = NULL) +
  scale_shape_manual(values = conv_shapes, name = NULL) +
  geom_hline(yintercept=0, lty='dashed', colour='grey70') +
  labs(title = "Final loss vs bias per guide",
       x = "Final ELBO loss", y = "bias (log2FC)") +
  theme_bw() +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```
