---
title: "bayesDREAM: Three-Step Illustration"
output:
  html_document:
    fig_width: 7
    fig_height: 5
    theme: minimal
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center")
library(ggplot2)
library(dplyr)
library(ggExtra)   # for marginal density panel
library(patchwork) # for plot layout (optional)

# ── Path to bayesDREAM output ──────────────────────────────────────────────
OUT_DIR <- file.path(dirname(rstudioapi::getSourceEditorContext()$path),
                     "output_illustration")
# Fallback if not running in RStudio:
if (!dir.exists(OUT_DIR)) OUT_DIR <- "output_illustration"

# ── Shared theme ───────────────────────────────────────────────────────────
theme_bd <- theme_classic(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    axis.title    = element_text(size = 12),
    legend.title  = element_text(size = 11),
    legend.text   = element_text(size = 10),
    legend.position = "right"
  )
```

---

## Step 1: Fit Technical

Technical variation shifts gene expression across cell lines. bayesDREAM estimates
cell-line–level offsets in NTC cells and corrects them before modelling cis effects.

```{r plot1-technical, fig.height=5, fig.width=6.5}
tech <- read.csv(file.path(OUT_DIR, "data_technical.csv"))

# ── Colour palette: grey reference + 3 purple shades ──────────────────────
group_levels <- c("Reference",
                  "Technical group A",
                  "Technical group B",
                  "Technical group C")
tech$group <- factor(tech$group, levels = group_levels)

group_colours <- c(
  "Reference"          = "#999999",
  "Technical group A"  = "#9B59B6",
  "Technical group B"  = "#6C3483",
  "Technical group C"  = "#D2B4DE"
)

# ── Group means ────────────────────────────────────────────────────────────
gm <- tech %>%
  group_by(group) %>%
  summarise(mx = mean(x_logcounts), my = mean(y_logcounts), .groups = "drop")

ref_mean <- gm %>% filter(group == "Reference")

# ── Arrow data: from each non-reference group mean → reference mean ────────
arrows_df <- gm %>%
  filter(group != "Reference") %>%
  mutate(
    xend = ref_mean$mx,
    yend = ref_mean$my
  )

# ── Plot ───────────────────────────────────────────────────────────────────
p1_tech <- ggplot(tech, aes(x = x_logcounts, y = y_logcounts, colour = group)) +
  geom_point(alpha = 0.55, size = 2) +
  geom_point(data = gm, aes(x = mx, y = my, colour = group),
             size = 5, shape = 18) +
  geom_segment(
    data = arrows_df,
    aes(x = mx, y = my, xend = xend, yend = yend, colour = group),
    arrow = arrow(length = unit(0.25, "cm"), type = "closed"),
    linewidth = 0.9,
    show.legend = FALSE
  ) +
  scale_colour_manual(values = group_colours, name = "Group") +
  labs(
    title = "Step 1: Fit Technical",
    x     = "Cis gene logcounts",
    y     = "Trans gene Y logcounts"
  ) +
  theme_classic() +
  theme(axis.text=element_blank(),
        axis.ticks=element_blank())
ggsave(
  filename = file.path("bayesDREAM_cartoon", "01_tech.png"),
  plot = p1_tech,
  width = 5,
  height = 3.2,
  dpi = 300,
  bg = "transparent",
  create.dir = TRUE
)
```

---

## Step 2: Fit Cis

bayesDREAM estimates guide-level true expression (*x*_true) for the cis gene,
integrating over cell-to-cell noise. Each dot is a cell; the vertical position
shows the posterior mean log₂(*x*_true) for that guide.

```{r plot2-cis, fig.height=5.5, fig.width=7}
library(stringr)
cis <- read.csv(file.path(OUT_DIR, "data_cis.csv"))
# x_true_log2fc already computed in Python relative to fitted NTC mean
cis$guide <- str_replace_all(cis$guide, 'KO', 'CI')
# ── Guide ordering (ascending x_true) and per-guide colours ──────────────
guide_order <- cis %>%
  group_by(guide) %>%
  summarise(mean_xt = mean(x_true_log2_mean, na.rm = TRUE), .groups = "drop") %>%
  arrange(mean_xt) %>%
  pull(guide)
cis$guide <- factor(cis$guide, levels = guide_order)

guide_colours <- c(
  "CI1"   = "#1A237E",
  "CI2"   = "#3949AB",
  "CI3"   = "#90CAF9",
  "NTC_1" = "#2ECC71",
  "CA1"   = "#EF9A9A",
  "CA2"   = "#C62828"
)

# ── Bonus plot: underlying x vs fitted x_true ─────────────────────────────
p_bonus <- ggplot(cis, aes(x = x_underlying_log2, y = x_true_log2_mean, colour = guide)) +
  geom_abline(intercept = 0, slope = 1, colour = "grey60", linetype = "dashed",
              linewidth = 0.5) +
  geom_jitter(alpha = 0.35, size = 1.6, width = 0.02, height = 0) +
  scale_colour_manual(values = guide_colours, name = "Guide") +
  labs(
    title = "Bonus: True x vs Fitted x_true",
    x     = expression(paste("True log"[2], " x")),
    y     = expression(paste("Fitted log"[2], " x"[true]))
  ) +
  theme_classic() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
ggsave(
  filename = file.path("bayesDREAM_cartoon", "00_bonus_xvsx.png"),
  plot = p_bonus,
  width = 5,
  height = 3.2,
  dpi = 300,
  bg = "transparent"
)

# ── Scatter: x = raw log₂ counts, y = fitted log₂ x_true (same scale) ────
p2_base <- ggplot(cis, aes(x = x_logcounts, y = x_true_log2_mean, colour = guide)) +
  geom_abline(intercept = 0, slope = 1, colour = "grey60", linetype = "dashed",
              linewidth = 0.5) +
  geom_jitter(alpha = 0.35, size = 1.6, width = 0.04, height = 0) +
  scale_colour_manual(values = guide_colours, name = "Guide") +
  labs(
    title = "Step 2: Fit Cis",
    x     = expression(paste("Cis gene log"[2], " counts")),
    y     = expression(paste("Fitted log"[2], " x"[true]))
  ) +
  theme_classic() +
  theme(axis.text=element_blank(),
        axis.ticks=element_blank())

# Add marginal density on the right (y-axis = fitted log₂ x_true)
p2_cis <- ggMarginal(p2_base, type = "density", margins = "y",
           groupColour = TRUE, groupFill = TRUE, alpha = 0.35)
ggsave(
  filename = file.path("bayesDREAM_cartoon", "02_cis.png"),
  plot = p2_cis,
  width = 5,
  height = 3.2,
  dpi = 300,
  bg = "transparent"
)
```

---

## Step 3: Fit Trans

bayesDREAM fits a non-monotonic (additive Hill) function linking cis gene
expression to trans gene expression. Each dot is a cell coloured by guide;
the curve shows the posterior mean fit with 90 % credible band.

```{r plot3-trans, fig.height=5, fig.width=7}
trans  <- read.csv(file.path(OUT_DIR, "data_trans.csv"))
hill   <- read.csv(file.path(OUT_DIR, "hill_curve.csv"))
refval <- read.csv(file.path(OUT_DIR, "ntc_reference.csv"))

x_ntc_log2 <- refval$x_ntc_log2[1]
y_ntc_log2  <- refval$y_ntc_log2[1]

# Convert y to log₂FC (relative to NTC y)
trans$y_log2fc      <- trans$y_logcounts   - y_ntc_log2
hill$y_post_mean_fc <- hill$y_post_mean    - y_ntc_log2
hill$y_post_lo_fc   <- hill$y_post_lo      - y_ntc_log2
hill$y_post_hi_fc   <- hill$y_post_hi      - y_ntc_log2

# Guide type factor ordering
type_levels <- c("CRISPRi", "NTC", "CRISPRa")
trans$guide_type <- factor(trans$guide_type, levels = type_levels)

p_trans <- ggplot() +
  # 90% credible band
  geom_ribbon(
    data = hill,
    aes(x = x_log2fc, ymin = y_post_lo_fc, ymax = y_post_hi_fc),
    fill = "#BDC3C7", alpha = 0.5
  ) +
  # Posterior mean Hill curve
  geom_line(
    data = hill,
    aes(x = x_log2fc, y = y_post_mean_fc),
    colour = "#2C3E50", linewidth = 1.1
  ) +
  # NTC reference lines (both at 0 in FC space)
  geom_vline(xintercept = 0, colour = "grey70", linetype = "dashed", linewidth = 0.6) +
  geom_hline(yintercept = 0, colour = "grey70", linetype = "dashed", linewidth = 0.6) +
  # Cell-level scatter
  geom_point(
    data = trans,
    aes(x = x_log2fc, y = y_log2fc, colour = guide_type),
    alpha = 0.45, size = 1
  ) +
  scale_colour_manual(
    values = c("CRISPRi" = "#3949AB", "NTC" = "#2ECC71", "CRISPRa" = "#C62828"),
    name = "Guide type"
  ) +
  scale_x_continuous(
    name   = expression(paste(log[2], "FC in cis gene")),
    breaks = 0, labels = "0",
    limits = c(-2.6, 1.6)
  ) +
  scale_y_continuous(
    name   = expression(paste(log[2], "FC in trans gene Y")),
    breaks = 0, labels = "0"
  ) +
  labs(title = "Step 3: Fit Trans") +
  theme_classic()
ggsave(
  filename = file.path("bayesDREAM_cartoon", "03_trans.png"),
  plot = p_trans,
  width = 5,
  height = 3.2,
  dpi = 300,
  bg = "transparent"
)
```
